\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-12-01 response_1/response_1.tex

4. p 10, `...relatively low overall with a mean estimate of N of 3.49 years (95\% HPD: 2.71-4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies...' I would not say that low Ne directly implies a ladder-like tree or vice versa, although it may produce such a tree in combination with highly dispersed sample times.

\textbf{We agree that casually calling the effective population size `low' does not explain the shape of the MERS-CoV phylogeny. We have rephrased this sentence to emphasise that the rate of coalescence observed in that particular part of the MERS-CoV phylogeny is high compared to the period over which sampling took place:}

\begin{quotation}
However, we do note that coalescence rate estimates are high relative to the sampling time period, with a mean estimate of $N_e\tau$ at 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder, as often seen in human influenza A virus phylogenies.
\end{quotation}

@@

\textbf{A very good point. We do not expect `sequencing capacity' to have been good in the initial stages of the outbreak and perhaps even to have peaked and declined at various stages, \textit{e.g.} during the Korean MERS outbreak. We think that addressing this analytically is perhaps a dimension too far, especially given the number of possible arrangements of sequencing intensity. We have added the following sentence to the results section:}

\begin{quotation}
Note that this ``sequencing capacity'' parameter does not vary over time, even though MERS-CoV sequencing efforts have varied in intensity, starting out slow due to lack of awareness, methods and materials and increasing in response to hospital outbreaks later.
\end{quotation}

6.3. The `model fitting' methods are ad-hoc. This is a situation where approximate Bayes computation would be appropriate. But I would try GEE/method of moments. I suspect the first three standardized moments of cluster sizes would be more useful than using mean/median/sd. - The moments could also be computed analytically if you dispense with the `sequencing capacity' aspect of the sampling model, so would be very fast.

\textbf{Our approach is indeed somewhat ad-hoc, though closely related to ABC.
We have essentially replaced the loss function in ABC with a slightly more complicated criterion based on  sequence data uncertainty.
Our approach is also equivalent to a fully Bayesian setting where the parameters of interest are assigned uniform priors.
Finally, while we completely agree with the reviewer that ABC would lend itself  very well to this estimation problem, it would also entail a good deal of calibration and experimentation, which we believe would be outside the scope of this paper.
We have followed the suggestion here to switch from mean/median/stdev to the first three standardised moments of sequence cluster size and found the results to be virtually identical.
This is to be expected, since the sample mean cluster size is the sufficient statistic for $R_0$ and the MLE for $k$ is proportional to the second (non-standardised) moment of the cluster sizes.
We do see somewhat better separation of simulations when plotting mean sequence cluster size versus their skewness in all figures summarising the results of Monte Carlo simulations.}

### 2017-12-01 response_1/response_1.tex

\textbf{A very good point. We do not expect `sequencing capacity' to have been good in the initial stages of the outbreak and perhaps even to have peaked and declined at various stages, \textit{e.g.} during the Korean MERS outbreak. We think that addressing this analytically is perhaps a dimension too far, especially given the number of possible arrangements of sequencing intensity. We have added the following sentence to the results section:}

\begin{quotation}
Note that this ``sequencing capacity'' parameter does not vary over time, even though MERS-CoV sequencing efforts have varied in intensity, starting out slow due to lack of awareness, methods and materials and increasing in response to hospital outbreaks later.
\end{quotation}

6.3. The `model fitting' methods are ad-hoc. This is a situation where approximate Bayes computation would be appropriate. But I would try GEE/method of moments. I suspect the first three standardized moments of cluster sizes would be more useful than using mean/median/sd. - The moments could also be computed analytically if you dispense with the `sequencing capacity' aspect of the sampling model, so would be very fast.

\textbf{Our approach is indeed somewhat ad-hoc, though closely related to ABC.
We have essentially replaced the loss function in ABC with a slightly more complicated criterion based on  sequence data uncertainty.
Our approach is also equivalent to a fully Bayesian setting where the parameters of interest are assigned uniform priors.
We have followed the suggestion here to switch from mean/median/stdev to the first three standardised moments of sequence cluster size and found the results to be virtually identical.
This is to be expected, since the sample mean cluster size is the sufficient statistic for $R_0$ and the MLE for $k$ is proportional to the second (non-standardised) moment of the cluster sizes.
We do see somewhat better separation of simulations when plotting mean sequence cluster size versus their skewness in all figures summarising the results of Monte Carlo simulations.}
Finally, while we completely agree with the reviewer that ABC would lend itself  very well to this estimation problem, it would also entail a good deal of calibration and experimentation, which we believe would be outside the scope of this paper.

### 2017-12-03 response_1/response_1.tex

\textbf{A very good point. We do not expect `sequencing capacity' to have been good in the initial stages of the outbreak and perhaps even to have peaked and declined at various stages, \textit{e.g.} during the Korean MERS outbreak. We think that addressing this analytically is perhaps a dimension too far, especially given the number of possible arrangements of sequencing intensity. We have added the following sentence to the results section:}

\begin{quotation}
Note that this ``sequencing capacity'' parameter does not vary over time, even though MERS-CoV sequencing efforts have varied in intensity, starting out slow due to lack of awareness, methods and materials and increasing in response to hospital outbreaks later.
\end{quotation}

6.3. The `model fitting' methods are ad-hoc. This is a situation where approximate Bayes computation would be appropriate. But I would try GEE/method of moments. I suspect the first three standardized moments of cluster sizes would be more useful than using mean/median/sd. - The moments could also be computed analytically if you dispense with the `sequencing capacity' aspect of the sampling model, so would be very fast.

\textbf{Our approach is indeed somewhat ad-hoc, though closely related to ABC.
We have essentially replaced the loss function in ABC with a slightly more complicated criterion based on  sequence data uncertainty.
Our approach is also equivalent to a fully Bayesian setting where the parameters of interest are assigned uniform priors.
We have followed the suggestion here to switch from mean/median/stdev to the first three standardised moments of sequence cluster size and found the results to be virtually identical.
This is to be expected, since the sample mean cluster size is the sufficient statistic for $R_0$ and the MLE for $k$ is proportional to the second (non-standardised) moment of the cluster sizes.
We do see somewhat better separation of simulations when plotting mean sequence cluster size versus their skewness in all figures summarising the results of Monte Carlo simulations.
Finally, while we completely agree with the reviewer that ABC would lend itself  very well to this estimation problem, it would also entail a good deal of calibration and experimentation, which we believe would be outside the scope of this paper. %LM: might want to re-order...
}

@@

\textbf{Apologies, we meant to say that $R_{0}$ is a more intuitive measure of epidemic behaviour than extrapolating risk from subtrees of a phylogeny. We have changed the sentence to say:}

\begin{quotation}
However, we wanted to translate this observation into an estimate of the basic reproductive number $R_{0}$ to provide an intuitive epidemic behaviour metric that does not require expertise in reading phylogenies.
\end{quotation}

5. Page 5. Your summary statistics are mean, median and standard deviation of cluster size. It seems odd not to have a statistic measuring skewness, given how skew the cluster sizes seem to be. Why did you not use one? Do you think you would have got similar results with one?

\textbf{Thank you for the comment.
% It was our thinking that mean, median and standard deviation of cluster sizes would be able to capture both the overall magnitude and skew of the empirical distribution.
The reason for choosing the sample mean, median 
We also accidentally omitted the number of clusters as a fourth parameter by which we filter simulated data.
In retrospect we should have been using skewness from the beginning. 
We have since fixed both these issues and now use mean, standard deviation, skewness and number of clusters as filters for simulated data throughout the manuscript.
This does not seem affect any of our results.}

### 2017-12-05 response_1/response_1.tex

6.4. Can you show a comparison of the empirical distribution of cluster sizes and the fitted distribution? A QQ plot? Are there any aspects of the empirical distribution that are not well characterised by the branching process model, eg at the tails?

\textbf{We have generated QQ plots of sequence cluster size distribution observed across the posterior distribution of structured coalescent trees (picked at random) against matching sequence cluster size distributions from the primary Monte Carlo simulation with bias parameter set to 2.0. Simulated sequence clusters tend to exhibit a left skew near the middle of the distribution, but not so much near the tails. We have added figure S9 to show this.}

\section*{Reviewer \#2}

1. Page 3 - where the authors state that ``Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitted between humans" some references should be added.

@@

\textbf{Apologies, we meant to say that $R_{0}$ is a more intuitive measure of epidemic behaviour than extrapolating risk from subtrees of a phylogeny. We have changed the sentence to say:}

\begin{quotation}
However, we wanted to translate this observation into an estimate of the basic reproductive number $R_{0}$ to provide an intuitive epidemic behaviour metric that does not require expertise in reading phylogenies.
\end{quotation}

5. Page 5. Your summary statistics are mean, median and standard deviation of cluster size. It seems odd not to have a statistic measuring skewness, given how skew the cluster sizes seem to be. Why did you not use one? Do you think you would have got similar results with one?

\textbf{Thank you for the comment.
% It was our thinking that mean, median and standard deviation of cluster sizes would be able to capture both the overall magnitude and skew of the empirical distribution.
The reason for choosing the sample mean, median and standard deviation was connected to the likelihood function we employed to model the distribution of cluster sizes.
As stated in the reply to Reviewer \#1's comment 6.3, the sample mean is the sufficient statistic for $R0$ and the MLE for the dispersion parameter depends on the sample mean and standard deviation only.
We also accidentally omitted the number of clusters as a fourth parameter by which we filter simulated data.
% In retrospect we should have been using skewness from the beginning. %LM: nah, I don't quite agree... Good idea in general, irrelevant here...
In the interest of investigating robustness, we now use mean, standard deviation, skewness and number of clusters as filters for simulated data throughout the manuscript.
This does not seem to affect any of our results.}

### 2017-12-05 mers-structure.tex

Structured coalescent approaches clearly show humans to be a terminal host for MERS-CoV, implying poor transmissibility.
However, we wanted to translate this observation into an estimate of the basic reproductive number $R_{0}$ to provide an intuitive epidemic behaviour metric that does not require expertise in reading phylogenies.
The parameter $R_{0}$ determines expected number of secondary cases in a single infections as well as the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}, Methods).
We estimate $R_{0}$ along with other relevant parameters via Monte Carlo simulation in two steps.
First, we simulate case counts across multiple outbreaks totaling 2000 cases using Equation \ref{clusterLikelihood} and then we subsample from each case cluster to simulate sequencing of a fraction of cases.
Sequencing simulations are performed via a multivariate hypergeometric distribution, where the probability of sequencing from a particular cluster depends on available sequencing capacity (number of trials), numbers of cases in the cluster (number of successes) and number of cases outside the cluster (number of failures).
In addition, each hypergeometric distribution used to simulate sequencing is concentrated via a bias parameter, that enriches for large sequence clusters at the expense of smaller ones (for its effects see Figure \ref{mers_epi_grid}).
This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from primary cases will be sampled exceedingly rarely.
We record the number, mean, standard deviation and skewness (third standard moment) of sequence cluster sizes in each simulation (left-hand panel in Figure \ref{mers_epi}) and extract the subset of Monte Carlo simulations in which these summary statistics fall within the 95\% highest posterior density observed in the empirical MERS-CoV data from structured coalescent analyses.
We record $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions), for these empirically matched simulations.
A schematic of this Monte Carlo procedure is shown in Figure \ref{mc_method}.
Since the total number of cases is fixed at 2000, higher $R_0$ results in fewer larger transmission clusters, while lower $R_0$ results in many smaller transmission clusters.

>>>>

Structured coalescent approaches clearly show humans to be a terminal host for MERS-CoV, implying poor transmissibility.
However, we wanted to translate this observation into an estimate of the basic reproductive number $R_{0}$ to provide an intuitive epidemic behaviour metric that does not require expertise in reading phylogenies.
The parameter $R_{0}$ determines expected number of secondary cases in a single infections as well as the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}, Methods).
We estimate $R_{0}$ along with other relevant parameters via Monte Carlo simulation in two steps.
First, we simulate case counts across multiple outbreaks totaling 2000 cases using Equation \ref{clusterLikelihood} and then we subsample from each case cluster to simulate sequencing of a fraction of cases.
Sequencing simulations are performed via a multivariate hypergeometric distribution, where the probability of sequencing from a particular cluster depends on available sequencing capacity (number of trials), numbers of cases in the cluster (number of successes) and number of cases outside the cluster (number of failures).
In addition, each hypergeometric distribution used to simulate sequencing is concentrated via a bias parameter, that enriches for large sequence clusters at the expense of smaller ones (for its effects see Figure \ref{mers_epi_grid}).
This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from primary cases will be sampled exceedingly rarely.
We record the number, mean, standard deviation and skewness (third standardised moment) of sequence cluster sizes in each simulation (left-hand panel in Figure \ref{mers_epi}) and extract the subset of Monte Carlo simulations in which these summary statistics fall within the 95\% highest posterior density observed in the empirical MERS-CoV data from structured coalescent analyses.
We record $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions), for these empirically matched simulations.
A schematic of this Monte Carlo procedure is shown in Figure \ref{mc_method}.
Since the total number of cases is fixed at 2000, higher $R_0$ results in fewer larger transmission clusters, while lower $R_0$ results in many smaller transmission clusters.
